---
title: Numerical integration in R
author: Philippe Massicotte
date: '2018-05-11'
slug: numerical-integration-in-r
categories:
  - R
  - Mathematics
tags:
  - R
subtitle: 'Math 101'
---

```{r setup, echo=FALSE, message=FALSE}
library(extrafont)
library(tidyverse)

## Set default ggplot2 font size and font familly
loadfonts(quiet = TRUE)
theme_set(theme_bw(base_size = 10, base_family = "IBM Plex Sans"))
```

Although R in not a symbolic language like [Mathematica](http://www.wolfram.com/mathematica/), [Maple](http://www.maplesoft.com/products/maple/) or [Matlab](http://www.mathworks.com/products/matlab/), it can be used for calculus operations. Let's get started. For the shake of simplicity, we will integrate the following equation:

$$
f(x) = \cos(x) + 2
$$

We first define the equation in the form of a function since it is a requirement to use the R [integrate](http://stat.ethz.ch/R-manual/R-devel/library/stats/html/integrate.html) function. Then, plot the function curve between 0 and 4.

```{r}
## Define the function we want to integrate cox(x) + 10
myfunc <- function(x) {
  cos(x) + 2
}

## Calculate the values
df <- tibble(x = seq(0, 4, 0.01), y = myfunc(x))

df %>%
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA))
```

Now, let's say we want to integrate between 1 and 3 which correspond to the gray area in the figure below.

```{r}
df %>%
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  geom_area(data = filter(df, between(x, 1, 3)))
```

Analytically, the solution would look like this:

$$
\begin{eqnarray}
 F(x)&=&\int\limits_1^3 \cos(x)+10 \,dx\\
  &=&\lbrack\sin(x)+2x\rbrack_1^3\\
  &=&[6 + \sin(3)] - [2 + \sin(1)]\\
  &=&4 + \sin(3) - \sin(1)\\
  &=&3.299649
\end{eqnarray}
$$

Now, let's integrate with R
```{r}
## Inegrate
integrate(myfunc, 1, 3)
```
The numerical and analytically solutions are consistent.

If you are aware of interesting package for numerical integration let me know!


