---
title: Analysis of speaker genders at the 2018 ArcticNet Annual Scientific Meeting
author: Philippe Massicotte
date: '2018-12-15'
tags:
  - Women in STEM
  - STEMWomen
  - ASM2018
  - ArcticNet
slug: gender-asm2018
---

```{r, setup, include=FALSE}
library(tidyverse)

theme_set(theme_bw(base_size = 12))
```

Last week I was participating to the [ArcticNet Annual Scientific Meeting 2018](http://www.arcticnetmeetings.ca/asm2018/) in Ottawa. I was pleasantly surprised to see that there was a large proportion of women present at the conference. After the panel discussion on *women in northern science* which took place on Thursday night, I decided to see if I could use R to scan the scientific PDF program to determine how many men and women were giving scientific presentations.

For this exercise, I will use the `pdftools` library to scan the program PDF.

```{r, cache=TRUE}
library(pdftools)

dat <- pdf_text("http://www.arcticnetmeetings.ca/asm2018/docs/PROGRAM_ASM2018_FINAL.pdf")
dat <- iconv(dat, from = "UTF-8", to = "ASCII//TRANSLIT")

res <- map(dat[18:51], function(x) { # Only scan pages 18-51

  # Use a simple regular expression to extract two consequtive upperletter
  # words. This could be better.
  # str_match_all(dat[[21]], "\\W+([A-Z\\s\\-]{1,})\\r\\n")
  ## Remove the \r if building on Linux
  str_match_all(x, "\\W+([[:upper:]]{3,}\\W+[[:upper:]]{3,})\\r\\n")[[1]][, 2]
  
}) %>%
  unlist() %>%
  tibble(full_name = .) %>%
  filter(!str_detect(full_name, "SESSION")) %>%
  separate(full_name, into = c("first_name", "last_name"))
  
```
If I did things correctly, there were `r nrow(res)` speakers at the conference.

```{r, echo=FALSE}
res
```

Then, using the `gender` package, we can determine the genders.

```{r}
library(gender)

res <- res %>% 
  nest(c(first_name, last_name)) %>%
  mutate(gender = map(data, ~ gender(.$first_name))) %>%
  unnest(gender) %>%
  mutate(gender = str_to_title(gender))

res

```

Finally, we can create a plot to visualize the data.

```{r}
res %>%
  count(gender) %>%
  ggplot(aes(x = gender, y = n)) +
  geom_col() +
  coord_flip() +
  theme(axis.title.y = element_blank()) +
  ylab("Number of speaker") +
  scale_y_continuous(breaks = seq(0, 200, by = 20)) +
  labs(title = "Number of male and female presenters at the ArcticNet 2018 meeting") +
  labs(caption = "Data source: http://www.arcticnetmeetings.ca/asm2018/docs/PROGRAM_ASM2018_FINAL.pdf")
```

This simple analysis confirms that there were as many men and women presenters at the conference. In the next graph we can see the most popular names of the speakers.

```{r}
res %>% 
  group_by(gender) %>% 
  count(name) %>% 
  arrange(desc(n)) %>% 
  slice(1:5) %>% 
  ggplot(aes(x = reorder(name, n, sort), y = n)) +
  geom_bar(stat = "identity") + 
  facet_wrap(~gender, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  ylab("Number of speakers") +
  labs(title = "Most frequent speaker names at the ArcticNet 2018 meeting") +
  theme(axis.title.y = element_blank())
```

